---
title: "Assignment 6 - SVMs and Unsupervised Learning"
author: "Slava Mikhaylov"
output: html_document
---

### Exercise 6.1

This problem involves the `OJ` data set which is part of the `ISLR` package.

(a) Create a training set containing a random sample of 800 observations, and a test set containing the remaining observations.

```{r}
library(ISLR)
set.seed(9004)
train <-  sample(dim(OJ)[1], 800)
OJ.train <-  OJ[train, ]
OJ.test <-  OJ[-train, ]
```

(b) Fit a support vector classifier to the training data using `cost=0.01`, with `Purchase` as the response and the other variables as predictors. Use the `summary()` function to produce summary statistics, and describe the results obtained.

```{r}
library(e1071)
svm.linear <-  svm(Purchase ~ . , kernel = "linear", 
                   data = OJ.train, cost = 0.01)
summary(svm.linear)
```

**Support vector classifier creates 432 support vectors out of 800 training points. Out of these, 217 belong to level `CH` and remaining 215 belong to level `MM`.**

(c) What are the training and test error rates?

```{r}
train.pred <-  predict(svm.linear, OJ.train)
table(OJ.train$Purchase, train.pred)
(82 + 53) / (439 + 53 + 82 + 226)
test.pred <-  predict(svm.linear, OJ.test)
table(OJ.test$Purchase, test.pred)
(19 + 29) / (142 + 19 + 29 + 80)
```

**The training error rate is 16.9% and test error rate is about 17.8%.**

(d) Use the `tune()` function to select an optimal cost. Consider values in the range 0.01 to 10.

```{r}
set.seed(1554)
tune.out <-  tune(svm, Purchase ~ . , data = OJ.train,
                  kernel = "linear", 
                  ranges = list(cost = 10^seq(-2, 1, by=0.25)))
summary(tune.out)
```

**Tuning shows that optimal cost is 0.3162.**

(e) Compute the training and test error rates using this new value for `cost`.

```{r}
svm.linear <-  svm(Purchase ~ . , kernel = "linear", 
                   data = OJ.train,
                   cost = tune.out$best.parameters$cost)
train.pred <-  predict(svm.linear, OJ.train)
table(OJ.train$Purchase, train.pred)
(57 + 71) / (435 + 57 + 71 + 237)
test.pred <-  predict(svm.linear, OJ.test)
table(OJ.test$Purchase, test.pred)
(29 + 20) / (141 + 20 + 29 + 80)
```

**The training error decreases to 16% but test error slightly increases to 18.1% by using best cost.**

(f) Repeat parts (b) through (e) using a support vector machine with a radial kernel. Use the default value for `gamma`.

```{r}
set.seed(410)
svm.radial <-  svm(Purchase ~ . , data = OJ.train, 
                   kernel = "radial")
summary(svm.radial)
train.pred <-  predict(svm.radial, OJ.train)
table(OJ.train$Purchase, train.pred)
(40 + 78) / (452 + 40 + 78 + 230)
test.pred <-  predict(svm.radial, OJ.test)
table(OJ.test$Purchase, test.pred)
(27 + 15) / (146 + 15 + 27 + 82)
```

**The radial basis kernel with default gamma creates 367 support vectors, out of which, 184 belong to level `CH` and remaining 183 belong to level `MM`. The classifier has a training error of 14.7% and a test error of 15.6% which is a slight improvement over linear kernel. We now use cross validation to find optimal gamma.**

```{r}
set.seed(755)
tune.out <-  tune(svm, Purchase ~ . , 
                  data = OJ.train, kernel = "radial",
                  ranges = list(cost = 10^seq(-2, 1, by=0.25)))
summary(tune.out)
svm.radial <-  svm(Purchase ~ . , data = OJ.train, 
                   kernel = "radial",
                   cost = tune.out$best.parameters$cost)
train.pred <-  predict(svm.radial, OJ.train)
table(OJ.train$Purchase, train.pred)
(77 + 40) / (452 + 40 + 77 + 231)
test.pred <-  predict(svm.radial, OJ.test)
table(OJ.test$Purchase, test.pred)
(28 + 15) / (146 + 15 + 28 + 81)
```

**Tuning slightly decreases training error to 14.6% and slightly increases test error to 16% which is still better than linear kernel.**

(g) Repeat parts (b) through (e) using a support vector machine with a polynomial kernel. Set `degree=2`.

```{r}
set.seed(8112)
svm.poly <-  svm(Purchase ~ . , data = OJ.train, 
                 kernel = "poly", degree = 2)
summary(svm.poly)
train.pred <-  predict(svm.poly, OJ.train)
table(OJ.train$Purchase, train.pred)
(32 + 105) / (460 + 32 + 105 + 203)
test.pred <-  predict(svm.poly, OJ.test)
table(OJ.test$Purchase, test.pred)
(12 + 37) / (149 + 12 + 37 + 72)
```

**Summary shows that polynomial kernel produces 452 support vectors, out of which, 232 belong to level `CH` and remaining 220 belong to level `MM`. This kernel produces a train error of 17.1% and a test error of 18.1% which are slightly higher than the errors produces by radial kernel but lower than the errors produced by linear kernel.**

```{r}
set.seed(322)
tune.out <-  tune(svm, Purchase ~ . , 
                  data = OJ.train, 
                  kernel = "poly", 
                  degree = 2, 
                  ranges = list(cost = 10^seq(-2, 1, by=0.25)))
summary(tune.out)
svm.poly <-  svm(Purchase ~ . , 
                 data = OJ.train, 
                 kernel = "poly", 
                 degree = 2, 
                 cost = tune.out$best.parameters$cost)
train.pred <-  predict(svm.poly, OJ.train)
table(OJ.train$Purchase, train.pred)
(37 + 84) / (455 + 37 + 84 + 224)
test.pred <-  predict(svm.poly, OJ.test)
table(OJ.test$Purchase, test.pred)
(13 + 34) / (148 + 13 + 34 + 75)
```

**Tuning reduces the training error to 15.12% and test error to 17.4% which is worse than radial kernel but slightly better than linear kernel.**

(h) Overall, which approach seems to give the best results on this data?

**Overall, radial basis kernel seems to be producing minimum misclassification error on both train and test data.**



### Exercise 6.2

In this problem, you will generate simulated data, and then perform PCA and K-means clustering on the data.

(a) Generate a simulated data set with 20 observations in each of three classes (i.e. 60 observations total), and 50 variables.
*Hint: There are a number of functions in `R` that you can use to generate data. One example is the `rnorm()` function; `runif()` is another option. Be sure to add a mean shift to the observations in each class so that there are three distinct classes.*

```{r}
set.seed(2)
x <-  matrix(rnorm(20*3*50, mean=0, sd=0.001), ncol=50)
x[1:20, 2] <-  1
x[21:40, 1] <-  2
x[21:40, 2] <-  2
x[41:60, 1] <-  1
```
**The concept here is to separate the three classes amongst two dimensions.**

(b) Perform PCA on the 60 observations and plot the first two principal component score vectors. Use a different color to indicate the observations in each of the three classes. If the three classes appear separated in this plot, then continue on to part (c). If not, then return to part (a) and modify the simulation so that there is greater separation between the three classes. Do not continue to part (c) until the three classes show at least some separation in the first two principal component score vectors.

```{r}
pca.out <-  prcomp(x)
summary(pca.out)
pca.out$x[,1:2]
plot(pca.out$x[,1:2], col=2:4, 
     xlab="Z1", ylab="Z2", pch=19) 
```

(c) Perform $K$-means clustering of the observations with $K = 3$. How well do the clusters that you obtained in $K$-means clustering compare to the true class labels?
*Hint: You can use the `table()` function in `R` to compare the true class labels to the class labels obtained by clustering. Be careful how you interpret the results: $K$-means clustering will arbitrarily number the clusters, so you cannot simply check whether the true class labels and clustering labels are the same.*

```{r}
km.out <-  kmeans(x, 3, nstart=20)
table(km.out$cluster, c(rep(1,20), rep(2,20), rep(3,20)))
```

**Perfect match.**

(d) Perform $K$-means clustering with $K = 2$. Describe your results.

```{r}
km.out <-  kmeans(x, 2, nstart=20)
km.out$cluster
```

**All of one previous class absorbed into a single class.**

(e) Now perform $K$-means clustering with $K = 4$, and describe your
results.

```{r}
km.out <-  kmeans(x, 4, nstart=20)
km.out$cluster
```

**All of one previous cluster split into two clusters.**

(f) Now perform $K$-means clustering with $K = 3$ on the first two principal component score vectors, rather than on the raw data. That is, perform $K$-means clustering on the $60 \times 2$ matrix of which the first column is the first principal component score vector, and the second column is the second principal component score vector. Comment on the results.

```{r}
km.out <-  kmeans(pca.out$x[,1:2], 3, nstart=20)
table(km.out$cluster, c(rep(1,20), rep(2,20), rep(3,20)))
```

**Perfect match, once again.**

(g) Using the `scale()` function, perform $K$-means clustering with $K = 3$ on the data after scaling each variable to have standard deviation one. How do these results compare to those obtained in (b)? Explain.

```{r}
km.out <-  kmeans(scale(x), 3, nstart=20)
km.out$cluster
```

**Poorer results than (b): the scaling of the observations effects the distance between them.**