---
title: "Assignment 8 - Tree-based Methods"
author: "Slava Mikhaylov"
output: html_document
---

Assignments for the course focus on practical aspects of the concepts covered in the lectures. Assignments are based on the material covered in James et al. You will start working on the assignment in the lab sessions after the lectures, but may need to finish them after class.

You will have two days to work on the assignments, submitting them via Moodle by 7pm on Thursday. We will subsequently open up solutions to the problem sets. 

### Exercise 8.1

In the lab, we applied random forests to the `Boston` data using $mtry=6$ and using $ntree=25$ and $ntree=500$. Create a plot displaying the test error resulting from random forests on this data set for a more comprehensive range of values for `mtry` and `ntree`. You can model your plot after Figure 8.10 in James et al.. Describe the results obtained.

**We will try a range of `ntree` from 1 to 500 and `mtry` taking typical values of $p$, $p/2$, $\sqrt{p}$. For Boston data, $p = 13$. We use an alternate call to `randomForest` which takes `xtest` and `ytest` as additional arguments and computes test MSE on-the-fly. Test MSE of all tree sizes can be obtained by accessing `mse` list member of `test` list member of the model.**

```{r}
library(MASS)
library(randomForest)
set.seed(1101)

# Construct the train and test matrices
train <-  sample(dim(Boston)[1], dim(Boston)[1] / 2)
X.train <-  Boston[train, -14]
X.test <-  Boston[-train, -14]
Y.train <-  Boston[train, 14]
Y.test <-  Boston[-train, 14]

p <-  dim(Boston)[2] - 1
p.2 <-  p / 2
p.sq <-  sqrt(p)

rf.boston.p <-  randomForest(X.train, Y.train, xtest=X.test,
                             ytest=Y.test, mtry=p, ntree=500)
rf.boston.p.2 <-  randomForest(X.train, Y.train, xtest=X.test,
                               ytest=Y.test, mtry=p.2, ntree=500)
rf.boston.p.sq <-  randomForest(X.train, Y.train, xtest=X.test,
                                ytest=Y.test, mtry=p.sq, ntree=500)

plot(1:500, rf.boston.p$test$mse, col="green", type="l", 
     xlab="Number of Trees", ylab="Test MSE", ylim=c(10, 19))
lines(1:500, rf.boston.p.2$test$mse, col="red", type="l")
lines(1:500, rf.boston.p.sq$test$mse, col="blue", type="l")
legend("topright", c("m=p", "m=p/2", "m=sqrt(p)"), 
       col=c("green", "red", "blue"), cex=1, lty=1)
```

**The plot shows that test MSE for single tree is quite high (around 18). It is reduced by adding more trees to the model and stabilizes around a few hundred trees. Test MSE for including all variables at split is slightly higher (around 11) as compared to both using half or square-root number of variables (both slightly less than 10).** 

### Exercise 8.2

Looking at the `Carseats` data from the `ISLR` package we will seek to predict `Sales` using regression trees and related approaches, treating the response as a quantitative variable.

(a) Split the data set into a training set and a test set.

```{r}
library(ISLR)
attach(Carseats)
set.seed(1)

train <-  sample(dim(Carseats)[1], dim(Carseats)[1] / 2)
Carseats.train <-  Carseats[train, ]
Carseats.test <-  Carseats[-train, ]
```

(b) Fit a regression tree to the training set. Plot the tree, and interpret the results. What test error rate do you obtain?

```{r}
library(tree)
tree.carseats <-  tree(Sales~ . , data=Carseats.train)
summary(tree.carseats)
plot(tree.carseats)
text(tree.carseats, pretty=0, cex=0.5)
pred.carseats <-  predict(tree.carseats, Carseats.test)
mean((Carseats.test$Sales - pred.carseats)^2)
```

**The test MSE is about $4.15$.**

(c) Use cross-validation in order to determine the optimal level of tree complexity. Does pruning the tree improve the test error rate?

```{r}
cv.carseats <-  cv.tree(tree.carseats, FUN=prune.tree)
par(mfrow=c(1, 2))
plot(cv.carseats$size, cv.carseats$dev, type="b")
plot(cv.carseats$k, cv.carseats$dev, type="b")

# Best size = 9
pruned.carseats <- prune.tree(tree.carseats, best=9)
par(mfrow=c(1, 1))
plot(pruned.carseats)
text(pruned.carseats, pretty=0, cex=0.5)
pred.pruned <-  predict(pruned.carseats, Carseats.test)
mean((Carseats.test$Sales - pred.pruned)^2)
```

**Pruning the tree in this case increases the test MSE to $4.99$.**

(d) Use the bagging approach in order to analyze this data. What test error rate do you obtain? Use the `importance()` function to determine which variables are most important.

```{r}
library(randomForest)
bag.carseats <-  randomForest(Sales~ . , data=Carseats.train, 
                              mtry=10, ntree=500, importance=TRUE)
bag.pred <-  predict(bag.carseats, Carseats.test)
mean((Carseats.test$Sales - bag.pred)^2)
importance(bag.carseats)
```

**Bagging improves the test MSE to $2.6$. We also see that `Price`, `ShelveLoc` and `Age` are three most important predictors of `Sale`.**

(e) Use random forests to analyze this data. What test error rate do you obtain? Use the `importance()` function to determine which variables are most important. Describe the effect of $m$, the number of variables considered at each split, on the error rate obtained.

```{r}
rf.carseats <-  randomForest(Sales ~ . , data=Carseats.train, 
                             mtry=5, ntree=500, importance=TRUE)
rf.pred <-  predict(rf.carseats, Carseats.test)
mean((Carseats.test$Sales - rf.pred)^2)
importance(rf.carseats)
```

**In this case, random forest worsens the MSE on test set to $2.87$. Changing $m$ varies test MSE between $2.6$ to $3$. We again see that `Price`, `ShelveLoc` and `Age` are three most important predictors of `Sale`.**

### Exercise 8.3

This problem involves the `OJ` dataset which is part of the `ISLR` package.

(a) Create a training set containing a random sample of 800 observations, and a test set containing the remaining observations.

```{r}
library(ISLR)
attach(OJ)
set.seed(1013)

train <-  sample(dim(OJ)[1], 800)
OJ.train <-  OJ[train, ]
OJ.test <-  OJ[-train, ]
```


(b) Fit a tree to the training data, with `Purchase` as the response and the other variables except for `Buy` as predictors. Use the `summary()` function to produce summary statistics about the tree, and describe the results obtained. What is the training error rate? How many terminal nodes does the tree have?

```{r}
library(tree)
oj.tree <-  tree(Purchase ~ . , data=OJ.train)
summary(oj.tree)
```

**The tree only uses two variables: `LoyalCH` and `PriceDiff`. It has $7$ terminal nodes. Training error rate (misclassification error) for the tree is $0.155$.**

(c) Type in the name of the tree object in order to get a detailed text output. Pick one of the terminal nodes, and interpret the information displayed.

```{r}
oj.tree
```

**Let's pick terminal node labeled "10)". The splitting variable at this node is `PriceDiff`. The splitting value of this node is $0.05$. There are $79$ points in the subtree below this node. The deviance for all points contained in region below this node is $80$. A * in the line denotes that this is in fact a terminal node. The prediction at this node is `Sales` = `MM`. About $19$% points in this node have `CH` as value of `Sales`. Remaining $81$% points have `MM` as value of `Sales`.**

(d) Create a plot of the tree, and interpret the results.

```{r 9d}
plot(oj.tree)
text(oj.tree, pretty=0, cex = 0.5)
```

**`LoyalCH` is the most important variable of the tree, in fact top 3 nodes contain `LoyalCH`. If `LoyalCH` < 0.27, the tree predicts `MM`. If `LoyalCH` > 0.76, the tree predicts `CH`. For intermediate values of `LoyalCH`, the decision also depends on the value of `PriceDiff`.**


(e) Predict the response on the test data, and produce a confusion matrix comparing the test labels to the predicted test labels. What is the test error rate?

```{r}
oj.pred <-  predict(oj.tree, OJ.test, type="class")
table(OJ.test$Purchase, oj.pred)
```

(f) Apply the `cv.tree()` function to the training set in order to determine the optimal tree size.

```{r}
cv.oj <-  cv.tree(oj.tree, FUN=prune.tree)
```

(g) Produce a plot with tree size on the $x$-axis and cross-validated classification error rate on the $y$-axis.

```{r}
plot(cv.oj$size, cv.oj$dev, 
     type="b", xlab="Tree Size", ylab="Deviance")
```

(h) Which tree size corresponds to the lowest cross-validated classification error rate?

**Size of 6 gives lowest cross-validation error.**

(i) Produce a pruned tree corresponding to the optimal tree size obtained using cross-validation. If cross-validation does not lead to selection of a pruned tree, then create a pruned tree with five terminal nodes.

```{r}
oj.pruned <-  prune.tree(oj.tree, best=6)
```

(j) Compare the training error rates between the pruned and unpruned trees. Which is higher?

```{r}
summary(oj.pruned)
```

**Misclassification error of pruned tree is exactly same as that of original tree --- $0.155$.**

(k) Compare the test error rates between the pruned and unpruned trees. Which is higher?

```{r}
pred.unpruned <-  predict(oj.tree, OJ.test, type="class")
misclass.unpruned <-  sum(OJ.test$Purchase != pred.unpruned)
misclass.unpruned / length(pred.unpruned)
pred.pruned <-  predict(oj.pruned, OJ.test, type="class")
misclass.pruned <-  sum(OJ.test$Purchase != pred.pruned)
misclass.pruned / length(pred.pruned)
```

**Pruned and unpruned trees have same test error rate of $0.189$.**

### Exercise 8.4

We now use boosting to predict `Salary` in the `Hitters` dataset, which is part of the `ISLR` package.

(a) Remove the observations for whom the salary information is unknown, and then log-transform the salaries.

```{r}
library(ISLR)
sum(is.na(Hitters$Salary))
Hitters <-  Hitters[-which(is.na(Hitters$Salary)), ]
sum(is.na(Hitters$Salary))
Hitters$Salary <-  log(Hitters$Salary)
```

(b) Create a training set consisting of the first 200 observations, and a test set consisting of the remaining observations.

```{r}
train <-  1:200
Hitters.train <-  Hitters[train, ]
Hitters.test <-  Hitters[-train, ]
```

(c) Perform boosting on the training set with 1,000 trees for a range of values of the shrinkage parameter $\lambda$. Produce a plot with different shrinkage values on the $x$-axis and the corresponding training set MSE on the $y$-axis.

```{r}
library(gbm)
set.seed(103)
pows <-  seq(-10, -0.2, by=0.1)
lambdas <-  10 ^ pows
length.lambdas <-  length(lambdas)
train.errors <-  rep(NA, length.lambdas)
test.errors <-  rep(NA, length.lambdas)

for (i in 1:length.lambdas) {
  boost.hitters <-  gbm(Salary ~ . , data=Hitters.train,
                        distribution="gaussian",
                        n.trees=1000,
                        shrinkage=lambdas[i])
  train.pred <-  predict(boost.hitters, Hitters.train, n.trees=1000)
  test.pred <-  predict(boost.hitters, Hitters.test, n.trees=1000)
  train.errors[i] <-  mean((Hitters.train$Salary - train.pred)^2)
  test.errors[i] <-  mean((Hitters.test$Salary - test.pred)^2)
}

plot(lambdas, train.errors, type="b", 
     xlab="Shrinkage", ylab="Train MSE", 
     col="blue", pch=20)
```

(d) Produce a plot with different shrinkage values on the $x$-axis and the corresponding test set MSE on the $y$-axis.

```{r}
plot(lambdas, test.errors, type="b", 
     xlab="Shrinkage", ylab="Test MSE", 
     col="red", pch=20)
min(test.errors)
lambdas[which.min(test.errors)]
```

**Minimum test error is obtained at $\lambda = 0.05$.**

(e) Compare the test MSE of boosting to the test MSE that results from applying two of the regression approaches seen in our discussions of regression models.

```{r}
lm.fit <-  lm(Salary ~ . , data=Hitters.train)
lm.pred <-  predict(lm.fit, Hitters.test)
mean((Hitters.test$Salary - lm.pred)^2)

library(glmnet)
set.seed(134)

x <-  model.matrix(Salary ~ . , data=Hitters.train)
y <-  Hitters.train$Salary
x.test <-  model.matrix(Salary ~ . , data=Hitters.test)
lasso.fit <-  glmnet(x, y, alpha=1)
lasso.pred <-  predict(lasso.fit, s=0.01, newx=x.test)
mean((Hitters.test$Salary - lasso.pred)^2)
```

**Both linear model and regularization like Lasso have higher test MSE than boosting.**

(f) Which variables appear to be the most important predictors in the boosted model?

```{r}
boost.best <-  gbm(Salary ~ . , data=Hitters.train,
                   distribution="gaussian", n.trees=1000,
                   shrinkage=lambdas[which.min(test.errors)])
summary(boost.best)
```

`CAtBat`, `CRBI` and `CWalks` are three most important variables in that order.

(g) Now apply bagging to the training set. What is the test set MSE for this approach?

```{r}
library(randomForest)
set.seed(21)
rf.hitters <-  randomForest(Salary ~ . , data=Hitters.train, 
                            ntree=500, mtry=19)
rf.pred <-  predict(rf.hitters, Hitters.test)
mean((Hitters.test$Salary - rf.pred)^2)
```

**Test MSE for bagging is about $0.23$, which is slightly lower than the best test MSE for boosting.**

### Exercise 8.5

This question uses the `Caravan` dataset, part of the `ISRL` package.

(a) Create a training set consisting of the first 1,000 observations, and a test set consisting of the remaining observations.

```{r}
library(ISLR)
train <-  1:1000
Caravan$Purchase <-  ifelse(Caravan$Purchase == "Yes", 1, 0)
Caravan.train <-  Caravan[train, ]
Caravan.test <-  Caravan[-train, ]
```

(b) Fit a boosting model to the training set with `Purchase` as the response and the other variables as predictors. Use 1,000 trees, and a shrinkage value of 0.01. Which predictors appear to be the most important?

```{r}
library(gbm)
set.seed(342)
boost.caravan <-  gbm(Purchase ~ . , 
                      data=Caravan.train, 
                      n.trees=1000, shrinkage=0.01,
                      distribution="bernoulli")
summary(boost.caravan)
```

**`PPERSAUT`, `MKOOPKLA` and `MOPLHOOG` are three most important variables in that order.**

(c) Use the boosting model to predict the response on the test data. Predict that a person will make a purchase if the estimated probability of purchase is greater than 20%. Form a confusion matrix. What fraction of the people predicted to make a purchase do in fact make one? How does this compare with the results obtained from applying KNN or logistic regression to this data set?

```{r}
boost.prob <-  predict(boost.caravan, Caravan.test, n.trees=1000,
                       type="response")
boost.pred <-  ifelse(boost.prob >0.2, 1, 0)
table(Caravan.test$Purchase, boost.pred)
34 / (137 + 34)
```

**About $20$% of people predicted to make purchase actually end up making one.**

```{r}
lm.caravan <-  glm(Purchase ~ . , data=Caravan.train, family=binomial)
lm.prob <-  predict(lm.caravan, Caravan.test, type="response")
lm.pred <-  ifelse(lm.prob > 0.2, 1, 0)
table(Caravan.test$Purchase, lm.pred)
58 / (350 + 58)
```

**About $14$% of people predicted to make purchase using logistic regression actually end up making one. This is lower than boosting.**

### Exercise 8.6 (Optional)

Apply boosting, bagging, and random forests to a data set of your choice. Be sure to fit the models on a training set and to evaluate their performance on a test set. How accurate are the results compared to simple methods like linear or logistic regression? Which of these approaches yields the best performance?

**In this exercise we examine the `Weekly` stock market data from the ISLR package.**

```{r}
set.seed(1)
library(ISLR)
summary(Weekly)
train <-  sample(nrow(Weekly), 2/3 * nrow(Weekly))
test <-  -train
```

**Logistic regression**

```{r}
glm.fit <-  glm(Direction ~ . -Year-Today, 
                data=Weekly[train,], 
                family="binomial")

glm.probs <-  predict(glm.fit, newdata=Weekly[test, ], 
                      type = "response")
glm.pred <-  rep("Down", length(glm.probs))
glm.pred[glm.probs > 0.5] <-  "Up"
table(glm.pred, Weekly$Direction[test])
mean(glm.pred != Weekly$Direction[test])
```

**Boosting**

```{r}
library(gbm)
Weekly$BinomialDirection <-  ifelse(Weekly$Direction == "Up", 1, 0)

boost.weekly <-  gbm(BinomialDirection~.-Year-Today-Direction,
                     data=Weekly[train,], 
                     distribution="bernoulli", 
                     n.trees=5000)

yhat.boost <-  predict(boost.weekly, 
                       newdata=Weekly[test,], 
                       n.trees=5000)

yhat.pred <-  rep(0, length(yhat.boost))
yhat.pred[yhat.boost > 0.5] <-  1
table(yhat.pred, Weekly$BinomialDirection[test])
mean(yhat.pred != Weekly$BinomialDirection[test])
```

**Bagging**

```{r}
Weekly <-  Weekly[,!(names(Weekly) %in% c("BinomialDirection"))]

library(randomForest)

bag.weekly <-  randomForest(Direction~.-Year-Today, 
                            data=Weekly, 
                            subset=train, 
                            mtry=6)

yhat.bag <-  predict(bag.weekly, newdata=Weekly[test,])
table(yhat.bag, Weekly$Direction[test])
mean(yhat.bag != Weekly$Direction[test])
```

**Random forests**

```{r}
rf.weekly <-  randomForest(Direction ~ . -Year-Today, 
                           data=Weekly, 
                           subset=train, 
                           mtry=2)

yhat.bag <-  predict(rf.weekly, newdata=Weekly[test,])
table(yhat.bag, Weekly$Direction[test])
mean(yhat.bag != Weekly$Direction[test])
```

**Best performance summary: Boosting resulted in the lowest validation set test error rate.**