---
title: "Assignment 5 - Resampling Methods"
author: "Slava Mikhaylov"
output: html_document
---

Assignments for the course focus on practical aspects of the concepts covered in the lectures. Assignments are based on the material covered in James et al. You will start working on the assignment in the lab sessions after the lectures, but may need to finish them after class.

You will have two days to work on the assignments, submitting them via Moodle by 7pm on Thursday. We will subsequently open up solutions to the problem sets. 

You will need to load the core library for the course textbook:
```{r}
library(ISLR)
```


## Exercise 5.1

Last week we used logistic regression to predict the probability of `default` using `income` and `balance` on the [Default](http://bit.ly/R_ISLR_Default) dataset. We will no estimate the test error of this logistic regression model using the validation set approach. Do not forget to set a random seed before beginning your analysis.

```{r}
library(ISLR)
summary(Default)
attach(Default)
```

 (a) Fit a logistic regression model that uses `income` and `balance` to predict `default`.

```{r}
set.seed(1)
glm.fit <-  glm(default ~ income + balance, 
                data = Default, 
                family = binomial)
```
   
 (b) Using the validation set approach, estimate the test error of this model. In order to do this, you must perform the following steps:
   
    i. Split the sample set into a training set and a validation set.
     
    ii. Fit a multiple logistic regression model using only the training observations.
     
    iii. Obtain a prediction of default status for each individual in the validation set by computing the posterior probability of default for that individual, and classifying the individual to the `default` category if the posterior probability is greater than 0.5.
     
    iv. Compute the validation set error, which is the fraction of the observations in the validation set that are misclassified.

```{r}
FiveB <-  function() {
# i.
train <- sample(dim(Default)[1], dim(Default)[1]/2)

# ii.
glm.fit <-  glm(default ~ income + balance, 
                data = Default, 
                family = binomial, 
                subset = train)

# iii.
glm.pred <- rep("No", dim(Default)[1]/2)
glm.probs <- predict(glm.fit, Default[-train,], type = "response")
glm.pred[glm.probs>.5] <-  "Yes"

# iv.
return(mean(glm.pred != Default[-train,]$default))
}

FiveB()
```

**2.86% test error rate from validation set approach.**
   
 (c) Repeat the process in (b) three times, using three different splits of the observations into a training set and a validation set. Comment on the results obtained.

```{r}
FiveB()
FiveB()
FiveB()
```

**It seems to average around 2.6% test error rate.**

 (d) Now consider a logistic regression model that predicts the probability of `default` using `income`, `balance`, and a dummy variable for `student`. Estimate the test error for this model using the validation set approach. Comment on whether or not including a dummy variable for `student` leads to a reduction in the test error rate.

```{r}
train <-  sample(dim(Default)[1], dim(Default)[1]/2)
glm.fit <-  glm(default ~ income + balance + student, 
                data = Default, 
                family = binomial, 
                subset=train)
glm.pred <-  rep("No", dim(Default)[1]/2)
glm.probs <-  predict(glm.fit, Default[-train,], type = "response")
glm.pred[glm.probs>.5] <-  "Yes"
mean(glm.pred != Default[-train,]$default)
```

**2.64% test error rate, with student dummy variable. Using the validation set approach, it doesn't appear adding the student dummy variable leads to a reduction in the test error rate.**


## Exercise 5.2

In the lab session for this week (Sections 5.3.2 and 5.3.3 in James et al.), we saw that the `cv.glm()` function can be used in order to compute the LOOCV test error estimate. Alternatively, one could compute those quantities using just the `glm()` and `predict.glm()` functions, and a `for` loop. You will now take this approach in order to compute the LOOCV error for a simple logistic regression model on the `Weekly` data set. Recall that in the context of classification problems, the LOOCV error is given in Section 5.1.5 (5.4, page 184).

```{r}
library(ISLR)
summary(Weekly)
set.seed(1)
attach(Weekly)
```

 (a) Fit a logistic regression model that predicts `Direction` using `Lag1` and `Lag2`.

```{r}
glm.fit <-  glm(Direction ~ Lag1 + Lag2, 
                data = Weekly, 
                family = binomial)
summary(glm.fit)
```

 (b) Fit a logistic regression model that predicts `Direction` using `Lag1` and `Lag2` using *all but the first observation*.

```{r}
glm.fit <-  glm(Direction ~ Lag1 + Lag2, 
                data = Weekly[-1,], 
                family = binomial)

summary(glm.fit)
```

 (c) Use the model from (b) to predict the direction of the first observation. You can do this by predicting that the first observation will go up if `P(Direction="Up"|Lag1, Lag2) > 0.5`. Was this observation correctly classified?

```{r}
predict.glm(glm.fit, Weekly[1,], type="response") > 0.5
```

**Prediction was UP, true Direction was DOWN.**

 (d) Write a `for` loop from i=1 to i=n, where n is the number of observations in the data set, that performs each of the following steps:

    i. Fit a logistic regression model using all but the i-th observation to predict `Direction` using `Lag1` and `Lag2`.
    
    ii. Compute the posterior probability of the market moving up for the i-th observation. 
    
    iii. Use the posterior probability for the i-th observation in order to predict whether or not the market moves up. 
    
    iv. Determine whether or not an error was made in predicting the direction for the i-th observation. If an error was made, then indicate this as a 1, and otherwise indicate it as a 0.

```{r}
count <-rep(0, dim(Weekly)[1])

for (i in 1:(dim(Weekly)[1])) {
   glm.fit <-  glm(Direction ~ Lag1 + Lag2, 
                   data = Weekly[-i,], 
                   family = binomial)
   
   is_up <-  predict.glm(glm.fit, Weekly[i,], type="response") > 0.5
   is_true_up <-  Weekly[i,]$Direction == "Up"
   if (is_up != is_true_up)
     count[i] <-  1
}

sum(count)
```

**490 errors.**

 (e) Take the average of the n numbers obtained in (d)iv in order to obtain the LOOCV estimate for the test error. Comment on the results.

```{r}
mean(count)
```

**LOOCV estimates a test error rate of 45%.** 

## Exercise 5.3

We will now perform cross-validation on a simulated data set.

 (a) Generate a simulated data set as follows:
   
```{r}
   set.seed(1)
   y <- rnorm(100)
   x <- rnorm(100)
   y <- x -2*x^2 + rnorm(100)
```

In this data set, what is $n$ and what is $p$? Write out the model used to generate the data in equation form.

**n = 100, p = 2.** 
**$Y = X - 2 X^2 + \epsilon$.**

 (b) Create a scatterplot of $X$ against $Y$. Comment on what you find.
 
```{r}
plot(x, y)
```

**Quadratic plot. $X$ from about -2 to 2. $Y$ from about -8 to 2.**

 (c) Set a random seed, and then compute the LOOCV errors that result from fitting the following four models using least squares:

    i. $Y = \beta_0 + \beta_1 X + \epsilon$
    ii. $Y = \beta_0 + \beta_1 X + \beta_2 X_2 + \epsilon$
    iii. $Y = \beta_0 + \beta_1 X + \beta_2 X_2 + \beta_3 X_3 + \epsilon$
    iv. $Y = \beta_0 + \beta_1 X + \beta_2 X_2 + \beta_3 X_3 + \beta_4 X_4 + \epsilon$.

Note you may find it helpful to use the `data.frame()` function
to create a single data set containing both $X$ and $Y$.

```{r}
library(boot)
Data <-  data.frame(x,y)
set.seed(1)

# i.
glm.fit <-  glm(y~x)
cv.glm(Data, glm.fit)$delta

# ii.
glm.fit <-  glm(y~poly(x,2))
cv.glm(Data, glm.fit)$delta

# iii.
glm.fit <-  glm(y~poly(x,3))
cv.glm(Data, glm.fit)$delta

# iv.
glm.fit <-  glm(y~poly(x,4))
cv.glm(Data, glm.fit)$delta
```


 (d) Repeat (c) using another random seed, and report your results. Are your results the same as what you got in (c)? Why?

```{r}
set.seed(10)

# i.
glm.fit <-  glm(y ~ x)
cv.glm(Data, glm.fit)$delta

# ii.
glm.fit <-  glm(y ~ poly(x,2))
cv.glm(Data, glm.fit)$delta

# iii.
glm.fit <-  glm(y ~ poly(x,3))
cv.glm(Data, glm.fit)$delta

# iv.
glm.fit <-  glm(y ~ poly(x,4))
cv.glm(Data, glm.fit)$delta
```

**Exact same, because LOOCV will be the same since it evaluates n folds of a single observation.**


 (e) Which of the models in (c) had the smallest LOOCV error? Is this what you expected? Explain your answer.

**The quadratic polynomial had the lowest LOOCV test error rate. This was expected because it matches the true form of $Y$.**

 (f) Comment on the statistical significance of the coefficient estimates that results from fitting each of the models in (c) using least squares. Do these results agree with the conclusions drawn based on the cross-validation results?
 
```{r}
summary(glm.fit)
```

**p-values show statistical significance of linear and quadratic terms, which agrees with the CV results.**
